E Coli Project Paper Incredibly Rough Draft

### 2013-08-27 paper/paper.tex
wrote some paper

%\pacs{61.20.Ne, 61.20.Gy, 61.20.Ja}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  This paper is about science.
\end{abstract}
\section{Introduction}
\fixme{I need a lot of references here!!!  (And the writing is just terrible)}
%It is vital that the process of bacterial cell division result in a single
nucleiod in each daughter cell.  The cell must avoid minicelling, or
splitting into daughter cells with lopsided volumes.  One of the mechanisms
the cell employs in order to avoid this is to allow a long FtsZ polymer
chain to develop on the cell wall in the center of the cell.  This chain
acts as a guide for the cell splitting process.  Huang et all have shown in
simulation that a system of Min protein interaction within the cell will
lead to a natural oscillation of the MinC protein, which exhibits an
aversion to the FtsZ polymer, from one cell end to the other that will
leave the center free of any build up of MinC.  This will allow for the
PtsZ chain to develop in the center and not at the ends, where the two
nucleiods are housed.  The interaction takes place between MinE,MinD, and
MinC, which naturally associates with MinD.

@@

\subsection{What is the MinD system and why is it important?}
\subsection{How proteins move in cell}
\section{Methods and Initial Conditions}
The model for the behavior of the MinD and MinE proteins inside the cell
implemented the same set of 5 reaction-diffusion equations described in the paper
by Huang et al (equations 1, 2, 3, 4, and 5). A 3d grid was constructed in
cartesian coordinates with a grid spacing of .05 $\mu$m. From there, we
were able to define a cell shape on the grid, and solve the
reaction-diffusion equations numerically to observe the time evolution of
the MinD and MinE concentrations inside the cell.

Kubitschek\cite{kubitschek1990cell} \cite{kubitschek1968linear} has
shown in multiple experiments that at the time of cell division cells
have a volume that is within a range of roughly $1\mu m^3$ to $2\mu
m^3$.  We follow Huang's simulations\cite{huang2003dynamic} and
Mannik's experiments \cite{mannik2012robustness} and model cells that
are slightly larger than this range.

We mean to study cells that are parrallel to those lodged into
crevices by Mannik so our flattened (non-pill shaped) cells have a
hieght of 0.25 \cite{mannick2012robustness} and appropriate
dimensions otherwise to yield the desired volume range.

Our simulation used the same diffusion constants and reaction rates as
Huang et al, which are
\begin{gather*} %format better
  \mathcal{D}_D = \mathcal{D}_{E}  = 2.5 \mu \textrm{m$^2$ / sec}, \\
  \sigma_D^{\textrm{ADP $\rightarrow$ ATP}}  = 1/\textrm{sec},  \sigma_D = 0.025 \mu \textrm{m/sec}, \\
  \sigma_{dD}  = 0.0015 \mu \textrm{m$^3$/sec}, \\
  \sigma_{de}  = 0.7/\textrm{sec}, \sigma_E = 0.093 \mu \textrm{m$^3$/sec}.
\end{gather*}
To test our computational model, we implemented a pill shaped cell, and
tested using the same cell parameters as Huang et al, which were a radius
of 0.5 $\mu$m in the middle and at the spherical endcaps, and two different
cell lengths of 4 $\mu$m and 10 $\mu$m. We found the same type of
oscillations as in their paper using these initial conditions, verifying
that our model works as intended. Below are snapshots of MinD and MinE
concentrations at 5 second timestamps in the 4 $\mu$m cell:
\newline
\newline
[insert 5 second time stamps of 4 $\mu$m sim]
\newline
\newline
We then began to define other, non-traditional cell shapes for the purpose of
modeling squished and perturbed E. coli cells, which were created
experimentally in Mannick et al. To achieve this, we went with a
cartesian lattice rather than the cylindrical lattice used in Huang et
al's simulations, as it allows for more flexibility in defining the
cell shape. Some of the cell shape models included a flattened pill
(stadium shape), an ellipsoid, a spherical cell, and various randomly
generated smooth shapes, such as those in the figures below.
\newline
\newline
[insert memf print of 2-3 cell shapes]
\newline
\newline
To interpret the results, we generated several different plot views of
the printed simulation data. These plots included a time averaged view
of the protein densities in the cell; a plot tracking the location of
protein concentrations that were global maxima in space and local maxima in
time; and an animated view that showed the actual dispersion of
protein concentrations in the cell over time.
\subsection{Mathematical Model}
\section{Specific Results}
\subsection{Pill Shape}

@@

\fixme{Be more exact about what exactly Nd is versus nATP, in terms of
  the unit and dimensions.  The plotting may need to be changed}
Figure %%\ref{frequency-plot-40-20-0-0-150} shows the protein density
fluctuations at a point in space adjecent to a polar wall.  At each
collection of peaks the proteins reach their zenith in an order that
agrees with the qualitative picture described by
Huang\cite{huang2003dynamic}, except for nADP peak.  The peaks start
with a maxima of ATP-MinD accumulating at the walls \fixme{Need a
  better way to refer to compounds}. This peak is followed by a peak
in Nde, or ATP-MinD-MinE, as the cell is converted on the wall from
the former to the latter compound.  As this compound splits apart and
leaves the wall we see a peak in nE, or MinE in the cytoplasm.  This
is then followed by a broader peak in nATP, or the MinD-ATP compound,
as minD-ADP is (changed but I forgot the name for it) into minD-ATP in
the cytoplasm.  The minD-ATP naturally diffuses away from the pole,
which is shown by the broad nature of this peak.  All of these peaks
fit the qualitative picture except for the sharp MinD-ADP peak.  One
could except that it is a sharp peak, meaning that the MinD-ADP proteins
do not last in the cytoplasm for very long before they are converted
(once again forgot the name) into MinD-ATP.  The location of the peak
in the time dimension is troubling, however, since that we would
expect it to occur just after the peak in Nde, since the wall-bound
MinD-ATP-MinE proteins are the ones that split up and leave the wall,
creating both the MinE and the MinD-ADP.  One would expect to see the
minD-ADP peak to coincide in time with the minE peak, but it instead
coincides with the wall-bound MinD-ATP-MinE peak. (Although you can
maybe sort of convince yourself otherwiase, really looking at it).

@@

To test our computational model, we implemented a pill shaped cell, and
tested using the same cell parameters as Huang et al, which were a radius
of 0.5 $\mu$m in the middle and at the spherical endcaps, and two different
cell lengths of 4 $\mu$m and 10 $\mu$m. We found the same type of
oscillations as in their paper using these initial conditions, verifying
that our model works as intended. Below are snapshots of MinD and MinE
concentrations at 5 second timestamps in the 4 $\mu$m cell:
\newline
\newline
[insert 5 second time stamps of 4 $\mu$m sim]
\newline
\newline
We then began to define other, non-traditional cell shapes for the purpose of
modeling squished and perturbed E. coli cells, which were created
experimentally in Mannick et al. To achieve this, we went with a
cartesian lattice rather than the cylindrical lattice used in Huang et
al's simulations, as it allows for more flexibility in defining the
cell shape. Some of the cell shape models included a flattened pill
(stadium shape), an ellipsoid, a spherical cell, and various randomly
generated smooth shapes, such as those in the figures below.
\newline
\newline
[insert memf print of 2-3 cell shapes]
\newline
\newline
To interpret the results, we generated several different plot views of
the printed simulation data. These plots included a time averaged view
of the protein densities in the cell; a plot tracking the location of
protein concentrations that were global maxima in space and local maxima in
time; and an animated view that showed the actual dispersion of
protein concentrations in the cell over time.
\section{Specific Results}
%-Datahjhuj and plots that show concrete results. Pill normal is for the establishing that we have what works, reference  other paper, then modify, is the idea.
\section{Pill Shape}
Pill Normal section -
Our goal with this project was to test whether or not the computational
model developed by Huang et al was consistent with the newer experimental
results (squishing E Coli) produced by Mannick et al. We generally see
that:

>>>>

%\pacs{61.20.Ne, 61.20.Gy, 61.20.Ja}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  This paper is about science.
\end{abstract}
\section{Introduction}
\fixme{I need a lot of references here!!!  (And the writing is just terrible)}
It is vital that the process of bacterial cell division result in a single
nucleiod in each daughter cell.  The cell must avoid minicelling, or
splitting into daughter cells with lopsided volumes.  One of the mechanisms
the cell employs in order to avoid this is to allow a long FtsZ polymer
chain to develop on the cell wall in the center of the cell.  This chain
acts as a guide for the cell splitting process.  Huang et all have shown in
simulation that a system of Min protein interaction within the cell will
lead to a natural oscillation of the MinC protein, which exhibits an
aversion to the FtsZ polymer, from one cell end to the other that will
leave the center free of any build up of MinC.  This will allow for the
PtsZ chain to develop in the center and not at the ends, where the two
nucleiods are housed.  The interaction takes place between MinE,MinD, and
MinC, which naturally associates with MinD.

@@

\subsection{What is the MinD system and why is it important?}
\subsection{How proteins move in cell}
\section{Methods and Initial Conditions}
The model for the behavior of the MinD and MinE proteins inside the
cell implements the same set of 5 reaction-diffusion equations studied in
the work of Huang et al (equations 1, 2, 3, 4, and
5)\cite{huang2003dynamic}. A 3d grid is constructed in cartesian
coordinates with a grid spacing of .05 $\mu$m. We define our cell
shapes and solve the reaction-diffusion equations numerically to
observe the time evolution of the MinD and MinE concentrations inside
the cell.

We start our cells with the MinD and MinE concentrations that are
reported as wild type concentrations by Huang et al . We as well use the same reaction
diffusion constants and and reaction rates,
\begin{gather*} %format better
  \mathcal{D}_D = \mathcal{D}_{E}  = 2.5 \mu \textrm{m$^2$ / sec}, \\
  \sigma_D^{\textrm{ADP $\rightarrow$ ATP}}  = 1/\textrm{sec},  \sigma_D = 0.025 \mu \textrm{m/sec}, \\
  \sigma_{dD}  = 0.0015 \mu \textrm{m$^3$/sec}, \\
  \sigma_{de}  = 0.7/\textrm{sec}, \sigma_E = 0.093 \mu \textrm{m$^3$/sec}.
\end{gather*}

We study cell shapes that are similar to those studied by Huang and
also new, flattened pancake-like shapes, inspired by the experimental work of
Mannik \cite{mannick2012robustness}.  Our pill shapes differ from
Huangs in that they are cylinders with hemisphere endcaps instead of
pure cylindridrical shapes.  Our cylinderidrical radius is $0.50\mu$
and the lengths of our cells (measured between the tips of the
endcaps) are $5\mu$, $4\mu$, $3\mu$, and $2.5\mu$.

With our flattened cell shapes we mean to study cells that are
parrallel to those lodged into crevices by Mannik so we give them
height of $0.25\mu$ \cite{mannick2012robustness}.  Viewed from the top
down the cells will have the shapes described below and viewed from
the side they have at their edges a semicircular protrusion (one may
imagine the edges of a pancake).

Our first group of (three) flattened cell shapes are triangular.  One
of them is equilateral, one has a 3-4-5 right angle shape, and the
last is iscoscolese.

Our second group of (four) flattened cells are our own creations,
designed to investigate different patterns of protein oscillation
behavoir.

The cell shapes and the sections that we split them into in our data
analysis are shown in Figure \fixme{We need a big figure with all of this!!}.

Kubitschek\cite{kubitschek1990cell} \cite{kubitschek1968linear} has
shown in multiple experiments that at the time of cell division cells
have a volume that is within a range of roughly $1\mu m^3$ to $2\mu
m^3$.  We follow Huang's simulations\cite{huang2003dynamic} and
Mannik's experiments \cite{mannik2012robustness} and model cells that
are slightly larger than this range.  Amoung our flattened cells the
two dimensional length scales are tuned so that every total cell
volume is very close to $3\mu^3$.  \fixme{should we just leave it to
  the reader to figure out for themselves what our pill shape volumes
  are?}

To interpret the results, we generated several different plot views of
the printed simulation data. These plots included a time averaged view
of the protein densities in the cell; a plot tracking the location of
protein concentrations that were global maxima in space and local maxima in
time; and an animated view that showed the actual dispersion of
protein concentrations in the cell over time.
\subsection{Mathematical Model}
\section{Specific Results}
\subsection{Pill Shape}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-p/plots/box-plot_D--p-400-50-0-0-1500}
  \includegraphics[width=\columnwidth]{../data/shape-p/plots/box-plot_D--p-200-50-0-0-1500}
  \caption{Total protein fluctuation in the right, middle, and left
    parts of a 5$\mu m$ by 1$\mu m$ pill shaped cell (above) and a
    3$\mu m$ by 1$\mu m$ pill shaped cell (below).  The vertical axis shows
    stacked the total number of proteins that are of four different
    compound stages and in the right section of the cell (bottom four
    colors), in the middle of the cell (middle four colors) and in the
    left section of the cell (top four colors). The different forms of
    protein changes while the total number of MinD protein in the cell
    remains constant, shown by the fact that the very top line is
    horizontal.}
  \label{total-oscillation-plot}
\end{figure}

Figure \ref{total-oscillation-plot} shows total protein
results for pill shaped cell of length $5\mu$ and radius $.5\mu$.  The
plot shows both the transerence between different stages of protein
compound and movement within the cell.  Starting at the beginning of a
period, we can see that first there is a spike on the left of MinD-ATP
attached to the wall that is accompanied by a growing number of
cytoplasmic MinD-ATP and MinD-ADP that are entering the left section,
followed by a high peak in MinD-MinE-ATP.  This peak gives way through
the middle to peaks on the right side of the cell, the first being a
small bump in cytoplasmic MinD-ATP that dissapears into a much larger
spike in wall attached MinD-ATP, which is followed by a peak in wall
attached MinD-MinE-ATP.  The pattern during each period clearly shows
the cycle of proteins rections, as well as the very regular nature of
the pole to pole protein oscillations.

\fixme{our 5 length cells are labeled as 4 cells in the file labels,
  since 4 is the length of the cylinder. In the text I'll refer to the
  5 length cells (file says 4) as 5 length cells.  What huang's paper
  calls 4 length cells will be labeled by our files as 3 length cells,
  but will be refered in our text as 4 length cells}.  We can see a
clear difference in period between the $5\mu$, $4\mu$, and $2\mu$
length cells.  The oscillation periods for our $4\mu$ show excellent
agreement with the simulations of Huang et all
\cite{huang2003dynamic}.  This is to be expected since we begin our
simulations with their reported wild type concentrations of MinD and
MinE proteins. The periods for the different pill lengths are shown in
the table below.  We can see a direct correltation between size and
period, as the proteins must diffuse further in a longer cell before
they accumulate on the opposite wall.
\newline
\newline
Period According to Pill Length
\begin{tabular}{|r|c|c|c|l|}
  \hline
  Length($\mu$) & 2.50 & 3.00 & 4.00 & 5.00\\
  \hline
  Period(sec) & sim & 33 & 38 & 48 \\ \cline{2-2}
  \hline
\end{tabular}

@@

To test our computational model, we implemented a pill shaped cell, and
tested using the same cell parameters as Huang et al, which were a radius
of 0.5 $\mu$m in the middle and at the spherical endcaps, and two different
cell lengths of 4 $\mu$m and 10 $\mu$m. We found the same type of
oscillations as in their paper using these initial conditions, verifying
that our model works as intended. Below are snapshots of MinD and MinE
concentrations at 5 second timestamps in the 4 $\mu$m cell:
\newline
\newline
[insert 5 second time stamps of 4 $\mu$m sim]
\newline
\newline
We then began to define other, non-traditional cell shapes for the purpose of
modeling squished and perturbed E. coli cells, which were created
experimentally in Mannick et al. To achieve this, we went with a
cartesian lattice rather than the cylindrical lattice used in Huang et
al's simulations, as it allows for more flexibility in defining the
cell shape. Some of the cell shape models included a flattened pill
(stadium shape), an ellipsoid, a spherical cell, and various randomly
generated smooth shapes, such as those in the figures below.
\newline
\newline
[insert memf print of 2-3 cell shapes]
\newline
\newline
To interpret the results, we generated several different plot views of
the printed simulation data. These plots included a time averaged view
of the protein densities in the cell; a plot tracking the location of
protein concentrations that were global maxima in space and local maxima in
time; and an animated view that showed the actual dispersion of
protein concentrations in the cell over time.
\section{Specific Results}
%-Datahjhuj and plots that show concrete results. Pill normal is for the establishing that we have what works, reference  other paper, then modify, is the idea.
\section{Pill Shape}
\fixme{Be more exact about what exactly Nd is versus nATP, in terms of
  the unit and dimensions.  The plotting may need to be changed}
Figure %%\ref{frequency-plot-40-20-0-0-150} shows the protein density
fluctuations at a point in space adjecent to a polar wall.  At each
collection of peaks the proteins reach their zenith in an order that
agrees with the qualitative picture described by
Huang\cite{huang2003dynamic}, except for nADP peak.  The peaks start
with a maxima of ATP-MinD accumulating at the walls \fixme{Need a
  better way to refer to compounds}. This peak is followed by a peak
in Nde, or ATP-MinD-MinE, as the cell is converted on the wall from
the former to the latter compound.  As this compound splits apart and
leaves the wall we see a peak in nE, or MinE in the cytoplasm.  This
is then followed by a broader peak in nATP, or the MinD-ATP compound,
as minD-ADP is (changed but I forgot the name for it) into minD-ATP in
the cytoplasm.  The minD-ATP naturally diffuses away from the pole,
which is shown by the broad nature of this peak.  All of these peaks
fit the qualitative picture except for the sharp MinD-ADP peak.  One
could except that it is a sharp peak, meaning that the MinD-ADP proteins
do not last in the cytoplasm for very long before they are converted
(once again forgot the name) into MinD-ATP.  The location of the peak
in the time dimension is troubling, however, since that we would
expect it to occur just after the peak in Nde, since the wall-bound
MinD-ATP-MinE proteins are the ones that split up and leave the wall,
creating both the MinE and the MinD-ADP.  One would expect to see the
minD-ADP peak to coincide in time with the minE peak, but it instead
coincides with the wall-bound MinD-ATP-MinE peak. (Although you can
maybe sort of convince yourself otherwiase, really looking at it).

Pill Normal section -
Our goal with this project was to test whether or not the computational
model developed by Huang et al was consistent with the newer experimental
results (squishing E Coli) produced by Mannick et al. We generally see
that:
